---
title: "QM model maps (raw)"
format: 
  html:
    toc: true
    toc-location: right
    toc-title: "Table of Contents"
    toc-depth: 2
    number-sections: true
    page-layout: full
    self-contained: true 
    collapsible: true
    collapse-level: 1
    grid:
      sidebar-width: 200px
      body-width: 880px
      margin-width: 200px
      gutter-width: 1.5rem
author: Gled Guri
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor: source
editor_options: 
  chunk_output_type: console
---

<style>
body {
  font-family: Arial, sans-serif;
  font-size: 18px;
}

h1 {
  font-size: 22px;  /* Adjust the size of H1 headers */
  font-weight: bold;  /* Make H1 headers not bold */
}

h2 {
  font-size: 18px;  /* Adjust the size of H2 headers */
  font-weight: normal;  /* Make H2 headers not bold */
}

h3 {
  font-size: 18px;  /* Adjust the size of H3 headers */
  font-weight: normal;  /* Make H3 headers not bold */
}

h4 {
  font-size: 18px;  /* Adjust the size of H4 headers */
  font-weight: normal;  /* Make H4 headers not bold */
}
</style>

```{r, message=FALSE, warning=FALSE}
library(here)
library(dplyr)
library(sf)
library(ggplot2)
library(rnaturalearth)
library(MoMAColors)

dat_sf <- readRDS(here('data','diff_spat_models','dat_sf.rds'))

coaststates <- ne_states(country='United States of America',returnclass = 'sf') %>% 
  filter(name %in% c('California','Oregon','Washington','Nevada')) %>% 
  st_transform(st_crs(dat_sf))
dat_bbox <- st_bbox(dat_sf)
```

```{r, include=TRUE, warning=FALSE}
#| fig-width: 13
#| fig-height: 13

plot_list_raw <- list()
sp_list <- dat_sf %>% 
  filter(species!='Merluccius productus') %>% 
  filter(species!='Hippoglossus stenolepis') %>% 
  pull(species) %>% 
  unique()

for (i in 1:length(sp_list)) {
  
plot_list_raw[[i]] <-
    dat_sf %>% filter(species==sp_list[i]) %>% 
  mutate(si_est=if_else(mean>-4,4,1)) %>% 
    ggplot() +
    geom_sf(aes(size = si_est, fill = mean), shape = 21, color = "black") +
    geom_sf(data = coaststates, fill = 'gray80') +
  ggtitle(paste0(sp_list[i])) +
    facet_wrap(~depth_cat, nrow = 1) +
    coord_sf(datum = NA) +
    scale_fill_gradientn(
      colors = moma.colors("ustwo", n = 21, direction = -1),
      limits = c(-4, 15),
      values = scales::rescale(c(-4:15)),
      name = "Log DNA conc") +
    scale_size_continuous(
    name = "P/A",
    limits = c(1,4),
    breaks = c(1, 4),
    labels = c('Absence', 'Presence'),
    range = c(1,4)
    ) +
    xlim(dat_bbox[1], dat_bbox[3]) +
    ylim(dat_bbox[2], dat_bbox[4]) +
    theme_bw()
ggsave(here('plots','QM posterior map',paste0(sp_list[i],'_raw.jpg')),width = 7,height = 6)
}

legend <- cowplot::get_legend(plot_list_raw[[1]]+theme(legend.justification = c(0,0.5)))
for (i in 1:length(sp_list)) {
plot_list[[i]] <- plot_list_raw[[i]]+theme(legend.position = "none")
}
```

```{r}
library(cowplot)
library(magick)

# Load the depth scheme graph
depth_scheme <- image_read(here('plots','QM posterior map','Literature_species_dist.jpeg'))
depth_scheme_plot <- ggdraw() + draw_image(depth_scheme)

# Create titles for columns
title_epipelagic <- ggdraw() + draw_label("Surface", fontface = 'bold')
title_mesopelagic <- ggdraw() + draw_label("Mid-water", fontface = 'bold')
title_bathypelagic <- ggdraw() + draw_label("Benthic", fontface = 'bold')

pp <-cowplot::plot_grid(
  cowplot::plot_grid(title_epipelagic, title_mesopelagic, title_bathypelagic, ncol = 3),
  
  cowplot::plot_grid(
    plot_list[[2]], plot_list[[7]], plot_list[[9]],
    plot_list[[5]], plot_list[[8]], plot_list[[4]],
    plot_list[[1]], plot_list[[10]], legend, 
    ncol = 3),nrow = 2, rel_heights = c(0.03, 1))

p <- cowplot::plot_grid(pp,depth_scheme_plot,nrow=1,rel_widths = c(10,7))
ggsave(here('plots','QM posterior map','Combined_plot_landscape.jpg'),p,width = 18,height = 12)

p <- cowplot::plot_grid(pp,depth_scheme_plot,nrow=2,rel_heights = c(10,5))
ggsave(here('plots','QM posterior map','Combined_plot_portrait.jpg'),p,width = 10,height = 16)
```