---
title: "Run_amp_eff_model_Mock_2"
format: 
  html:
    toc: true
    toc-location: left
    toc-title: "Table of Contents"
    toc-depth: 2
    number-sections: true
    page-layout: full
    self-contained: true 
    collapsible: true
    collapse-level: 1
    # code-summary: "Show/Hide Code"  # Toggle text for code folding
    # css: 'style_gg.css'
    grid:
      sidebar-width: 200px
      body-width: 880px
      margin-width: 200px
      gutter-width: 1.5rem
author: Gled Guri
date: "`r format(Sys.time(), '%d %B, %Y')`"
editor: visual
editor_options: 
  chunk_output_type: console
---

<style>
body {
  font-family: Arial, sans-serif;
  font-size: 18px;
}

h1 {
  font-size: 22px;  /* Adjust the size of H1 headers */
  font-weight: normal;  /* Make H1 headers not bold */
}

h2 {
  font-size: 18px;  /* Adjust the size of H2 headers */
  font-weight: normal;  /* Make H2 headers not bold */
}

h3 {
  font-size: 18px;  /* Adjust the size of H3 headers */
  font-weight: normal;  /* Make H3 headers not bold */
}

h4 {
  font-size: 18px;  /* Adjust the size of H4 headers */
  font-weight: normal;  /* Make H4 headers not bold */
}
</style>

```{r, include=FALSE}
dwidth <- 150
options(width = dwidth)
```

```{r, include=FALSE, warning=FALSE, message=FALSE}
options(width = 300) #wd
```

```{r, include=FALSE, warning=FALSE, message=FALSE}
options(width = dwidth)#wd
```

# Data prep

## Load libraries

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
library(tibble)
library(here)
library(tidyr)
library(rstan);options(mc.cores = parallel::detectCores())
library(QM);#devtools::install_github("gledguri/QM",force = T)
library(MoMAColors);#devtools::install_github("BlakeRMills/MoMAColors");require(MoMAColors)
library(ggu.base.fun);#devtools::install_github('gledguri/ggednasd/ggu.base.fun',force = T)
```

## Load Data

```{r, warning=FALSE, message=FALSE}
mock_ini_prop <- read.csv(here('data/mocks/Mock_v2_v3_v4/Data/','expected_props_corrected_oct_24.csv'))
mb <- read.csv(here('data/mocks/Mock_v2_v3_v4/Data/','5_ASV_MFU_322_GG_annotated_collapsed_reitterated.csv'))
```

## Curate data

```{r, warning=FALSE, message=FALSE}
#Pivot wider mock initial data
mock_ini_prop_tr <- mock_ini_prop  %>% 
	mutate(x=paste0(mock,'_',even_skew,'_',prop_type)) %>% select(-mock,-even_skew,-prop_type)

mock_ini_prop_tr <-
	inv.trans(
	input.X = mock_ini_prop_tr$BestTaxon,
	input.Y = mock_ini_prop_tr$x,
	input.Z = mock_ini_prop_tr$Prop)

#Show mock initial proportion data
mock_ini_prop_tr %>% rownames_to_column() %>% 
	as_tibble()

#Shorten column names of MB reads output
colnames(mb) <- 
	colnames(mb) %>% as.data.frame() %>% setNames('c') %>% 
	mutate(c=gsub('MFU.CPS.','',c)) %>% 
	mutate(c=gsub('d1.','',c)) %>% 
	mutate(c=gsub('mock','M',c)) %>% 
	mutate(c=gsub('skew','s',c)) %>% 
	mutate(c=gsub('even','e',c)) %>% 
	mutate(c=gsub('mtDNA','mt',c)) %>% 
	mutate(c=gsub('gDNA','g',c)) %>% 
	pull(c)

# Select only the samples from Mock 2,3,and 4
mock_mb <-
	mb %>% 
	select(Species_name,
				 colnames(.)[grepl("^M([234])\\.([es])\\.",colnames(.))]) %>% 
	mutate(Species_name= if_else(is.na(Species_name),'Unknown',Species_name))%>% 
	column_to_rownames('Species_name') %>% 
	filter(rowSums(.)>0)
```

## Curate data 2

```{r Combine MB reads with initial prop, warning=FALSE, message=FALSE}
# Combine MB reads with initial prop 
#combine mock 2 data (& selecting only the mtDNA)
mock_2_ini_prop_tr <-
	mock_ini_prop_tr %>% 
	select(colnames(.)[grepl('mock2',colnames(.))]) %>% 
	select(colnames(.)[grepl('mtDNA',colnames(.))]) %>% 
	filter(rowSums(.)>0)

mock_2_mb <- mock_mb %>%
	select(colnames(.)[grepl('M2.',colnames(.))])

#Combine MB reads with initial prop 
mock2 <- custom_cbind(mock_2_ini_prop_tr,mock_2_mb)

#Show the combined data (ini_prop and MB reads)
mock2 %>% rownames_to_column() %>% 
	as_tibble() %>% print(n=nrow(.))

#Show species that are not inputed in mocks but have MB reads
mock2 %>% filter(is.na(mock2_even_mtDNA)) %>% 
	rownames_to_column() %>% 
	as_tibble() %>% print(n=nrow(.))
```

## Filter sp

```{r, warning=FALSE, message=FALSE}
#'Filter those species out (considering they're artifacts)
#'fixing the proportion for the remaining species that are actually 
#'in initial mocks and MB outputs
	mock2 <- mock2 %>% 
  filter(!apply(., 1, function(row) any(is.na(row)))) %>% 
	mutate(mock2_even_mtDNA=mock2_even_mtDNA/sum(mock2_even_mtDNA)) %>% 
	mutate(mock2_skew_mtDNA=mock2_skew_mtDNA/sum(mock2_skew_mtDNA))
```

## Curate data 3

```{r, warning=FALSE, message=FALSE}
#Select the reference species and create species_idx
mock2 <- mock2 %>% rownames_to_column('Species') %>% 
		mutate(Species=if_else(Species=='Merluccius productus',
											'Zz_Merluccius productus',
											Species)) %>% 
		arrange(Species) %>% 
		mutate(sp_idx=c(1:nrow(.)))
```

```{r, warning=FALSE, message=FALSE}
#Split the mock df into even and skewed (2 data frames)
mock2_even <- mock2 %>% 
	select(Species,
		colnames(.)[grepl('even',colnames(.))],
		colnames(.)[grepl('M2.e',colnames(.))],
		sp_idx)

#Show mock 2 even
mock2_even %>% as_tibble()

mock2_skew <- mock2 %>% 
	select(Species,
					 colnames(.)[grepl('skew',colnames(.))],
					 colnames(.)[grepl('M2.s',colnames(.))],
				 sp_idx)

#Show mock 2 skew
mock2_skew %>% as_tibble()
```

# Model

## Load QM model

```{r, warning=FALSE, message=FALSE, results='hide'}
M3_2_stan_code <- 
"
data {
	// Intigers
  int N_sp_i_MC2; // Number of species
  int N_obs_Y_M5; // Number of samples included in mock community
  //
  // Data
  vector[N_sp_i_MC2] alr_M5; // Additive log(e)-ratio of initial concentration of mock community
  array[N_sp_i_MC2, N_obs_Y_M5] int Y_M5; // Sequence reads of mock community samples
  int NPCR;  // Number of PCR reactions
  //
  // Parameters
  real alpha_magnitude; // dispersion parameter of alpha
}
parameters {
  vector[N_sp_i_MC2-1] alpha_raw;
}
transformed parameters{
  matrix[N_sp_i_MC2,N_obs_Y_M5] gamma_M5;
  matrix[N_sp_i_MC2,N_obs_Y_M5] psi_M5;
  vector[N_sp_i_MC2] alpha;
  alpha[1:(N_sp_i_MC2-1)] = alpha_raw * alpha_magnitude;
  alpha[N_sp_i_MC2] = 0;
  for (i in 1:N_obs_Y_M5){
    for(j in 1:N_sp_i_MC2){
      gamma_M5[j,i] = alr_M5[j]+(NPCR*(alpha[j]));
    }
  }
  for (i in 1:N_obs_Y_M5){
    psi_M5[,i] = softmax(gamma_M5[,i]);
  }
}
model {
  for (i in 1:N_obs_Y_M5){
    Y_M5[,i] ~ multinomial(psi_M5[,i]);
  }
  alpha_raw ~ std_normal();
}
"

M3_2 <- stan_model(model_code = M3_2_stan_code,model_name = 'M3_2 model')
```

## Prep stan_data

```{r, warning=FALSE, message=FALSE}
#Preparing stan for mock 2 even
stan_data_mock_2_even <- prep_stan_M3_2(
	data = mock2_even,
	mock_sequencing_columns = c("M2.e.1_S1","M2.e.2_S2","M2.e.3_S3"),
	mock_initial_concentration = 'mock2_even_mtDNA',
	species_index = 'sp_idx',
	mock_species_names = 'Species',
	number_of_PCR = 43,
	alpha_magnitude = 0.001)

#Preparing stan for mock 2 skew
stan_data_mock_2_skew <- prep_stan_M3_2(
	data = mock2_skew,
	mock_sequencing_columns = c("M2.s.1_S9","M2.s.2_S10","M2.s.3_S11"),
	mock_initial_concentration = 'mock2_skew_mtDNA',
	species_index = 'sp_idx',
	mock_species_names = 'Species',
	number_of_PCR = 43,
	alpha_magnitude = 0.001)
```

## Run stan

```{r, warning=FALSE, message=FALSE,results='hide'}
#Run alpha model mock 2 even
stanMod_3_mock_2_even <- sampling(
	object = M3_2,
	data = stan_data_mock_2_even,
	iter = 2000,
	warmup = 1000,
	chains = 4)

#Run alpha model mock 2 skew
stanMod_3_mock_2_skew <- sampling(
	object = M3_2,
	data = stan_data_mock_2_skew,
	iter = 2000,
	warmup = 1000,
	chains = 4)
```

# Output

## Extract amplification efficiencies

```{r, warning=FALSE, message=FALSE}
#mock 2 even
amp_eff_param_mock_2_even <-
	stan_data_mock_2_even$Species %>%
	cbind(.,
				amp_eff_param_extract(stanMod_3_mock_2_even))

amp_eff_param_mock_2_even %>% as_tibble()

#mock 2 skew
amp_eff_param_mock_2_skew <-
	stan_data_mock_2_skew$Species %>%
	cbind(.,
				amp_eff_param_extract(stanMod_3_mock_2_skew))

amp_eff_param_mock_2_skew %>% as_tibble()
```

## Curate data 4

```{r, warning=FALSE, message=FALSE}
amp_eff_param_mock_2 <-
	cbind(amp_eff_param_mock_2_even %>%
					select(Species,mean) %>%
					rename(M2_even='mean'),
				amp_eff_param_mock_2_skew %>%
					select(mean) %>%
					rename(M2_skew='mean')) %>%
	pivot_longer(cols = starts_with("M2_"),
							 names_to = "mock_type",
							 values_to = "alpha")
```

```{r,warning=FALSE, message=FALSE}
# saveRDS(mock2_even,here('data/mocks/Mock_v2_v3_v4/Output/','mock2_even_filtered.rds'))
# saveRDS(mock2_skew,here('data/mocks/Mock_v2_v3_v4/Output/','mock2_skew_filtered.rds'))
# saveRDS(amp_eff_param_mock_2,here('data/mocks/Mock_v2_v3_v4/Output/','amp_eff_param_mock_2_all_sp.rds'))
```

## Plot alphas

```{r, warning=FALSE, message=FALSE}
#mock 2 (even vs skew)
ggplot(data = amp_eff_param_mock_2)+
	geom_point(aes(y=Species,x=alpha,color=mock_type))+
	ggtitle('Mock 2 even vs skew')+
	scale_color_manual(values = moma.colors("Lupi",n = le(amp_eff_param_mock_2$mock_type)))+
	# xlim(-0.06,0.06)+
	theme_bw()+
	geom_vline(xintercept = 0,
						 linetype = "dashed",
						 color = 'black',
						 linewidth=0.2)+
	labs(x = "Alpha", y = "Species")

#previously problematic species Trachurus symmetricus is actually being correctly
#estimated the same between even and skew
amp_eff_param_mock_2 %>% filter(grepl('Trachurus symmetricus',Species))
```
